Coherent IR-Hyper-Raman Four Wave Mixing Vibrational Spectroscopy

### 2024-07-29 paper.tex
last few things

\section{Introduction}
Coherent multidimensional spectroscopy (CMDS) is a family of nonlinear spectroscopy methods that form the optical analogue of multidimensional nuclear magnetic resonance (NMR) spectroscopy.\cite{Cho2008}
Multiresonant, four wave mixing CMDS experiments, first proposed by Oudar and Shen in 1980,\cite{RN307} directly probe coupling and correlations between different vibrational, electronic, and vibronic states. \cite{RN281, RN103, Cho2008} 
Four wave mixing CMDS has resolved anharmonicities, ultrafast dynamics, and other inter- and intra-molecular couplings in numerous systems. \cite{Cho2008, Gaynor2017, Ziegler2018, Ogilvie2019, Bonn2021, RN325, Chen2024}
To target vibrational and electronic coupling, methods commonly make use of Raman transitions, e.g., doubly vibrationally enhanced spectroscopy (DOVE), triple sum frequency (TSF), coherent anti-Stokes Raman spectroscopy (CARS), resonant impulse stimulated Raman spectroscopy (RISRS).\cite{RN425, Dhar1994, RN345, Cho2000, RN491}
Here we explore the less-utilized hyper-Raman transitions in four wave mixing CMDS.

@@

\autoref{fig:comparisonwmel} shows both HDFG and HSFG processes and compares them with related spectroscopies that investigate vibrational-electronic coupling.
The three-wave mixing ($\chi^{(2)}$) techniques are often less applicable as they require non-centrosymmetry.\cite{RN227}
Hyper-Raman scattering itself is a six-wave mixing ($\chi^{(5)}$) technique; stimulated hyper-Raman will be weak and susceptible to FWM cascades.\cite{RN515, RN243, Cho2000_Cascade}
A coherent six-wave mixing analogue of hyper-Raman scattering, coherent anti-Stokes hyper-Raman spectroscopy (CAHRS), has been proposed and explored theoretically by several authors; however, it is also highly susceptible to FWM cascades.\cite{Berger1978, Bjarnason1980, RN243, Cho1997, Cho1998, Cho2000_Cascade}
Of the many four-wave mixing techniques, HDFG and HSFG have strong similarity to Raman, stimulated Raman, and RISRS in that the electronic state must remove or add a $v$ quantum to the ground state (``ev'' coupling).
With the other four-wave mixing techniques (\autoref{fig:comparisonwmel}, second row), the output signal depends on more states than just a single vibrational mode and electronic state (``evx'' coupling). 
As a result, the relationship between the output and vibrational-electronic coupling is more detailed, but also more complicated.
HDFG, HSFG and Raman are more direct probes of electronic-vibrational coupling.
Notably, HDFG and HSFG can be implemented as 2-color experiments (with $\omega_2=\omega_3$); vibrational sum frequency generation (vSFG) or difference frequency generation (vDFG) setups can be trivially reconfigured to perform HSFG or HDFG measurements, respectively, to investigate ev coupling in bulk systems.

@@

Inspired by recent work that demonstrated the presence of HDFG in ultrafast DOVE and terahertz-IR-visible (TIRV) experiments, we investigate the parameters which result in HDFG output. \cite{Cho2000, Bonn2024, McDonnell2024}
This paper is organized as follows.
In \autoref{steadystate}, we first identify the selection rules that drives HDFG in the absence of electronic resonance. 
The selection rules of electronically-resonant HDFG are then identified through a Herzberg-Teller dipole expansion.
HDFG is found to be allowed in any harmonic system that has infrared active vibrations.
A simple harmonic oscillator model system is used to simulate HDFG spectra and illustrate how the excited state potential energy surface affects the hyper-Raman excitation spectrum.
After developing selection rules, the site selective properties, useful experimental aspects of HDFG, and a new scheme to extract the hyper-Raman hyperpolarizability ($\beta_{ijk}$) from HDFG spectra are discussed in \autoref{quant}.
The main findings are summarized in \autoref{conclusion}.

@@

It is useful to expose relationships between transition dipoles and hyper-Raman hyperpolarizabilities in the driven limit. \cite{Simpson2004}
Under the electric dipole approximation, the I$^{th}$ component of the third order nonlinear output polarization, ${P}^{(3)}_I$, of a four wave mixing process, induced by electric fields $E_J$, $E_K$, and $E_L$ at output frequency $\omega_4=-\omega_1 + \omega_2 + \omega_3$ is written as (using Einstein summation) \cite{RN307}
\begin{equation} \label{polarization}
{P}^{(3)}_I (\omega_4)  = \chi^{(3)}_{IJKL} E_J(\omega_3) E_K(\omega_2) E_L(\omega_1) 
\end{equation}
where $\chi^{(3)}_{IJKL}$ is the IJKL element of the third order electrical susceptibility, a rank four tensor, generally written as
\begin{equation}
	\chi^{(3)}_{IJKL} = NF(\omega_4) \langle \gamma_{ijkl} \rangle
\end{equation}
where $N$ is a number density, $F$ is the Lorentz local field factor, and $\gamma_{ijkl}$ is the third order polarizability (i.e., second hyperpolarizability). 
The brackets indicate an orientational (or rotational) average.\cite{Andrews1977}
Uppercase letters refer to laboratory frame coordinates and lower case letters refer to corresponding molecular frame coordinates.

@@

When $\omega_2 + \omega_3$ become electronically resonant, the interference changes for the \autoref{fig:sivewmel2}c-f pathways.
Note that only pathways (c,e,f) possess identical transition moments.
As a result, pathways (c,d) survive and can provide output when $\omega_2+\omega_3$ is resonant with an electronic state. 
These results provide a simple rule of thumb for collecting HDFG spectra: use \autoref{fig:sivewmel2}a,b pathways to measure spectra when significantly detuned from an electronic resonance, and any pathway can measure spectra when electronically resonant.
However, since electronic states commonly have short dephasing times and have significantly inhomogeneous broadening,\cite{Dong2015, Lewis2015} it is likely the \autoref{fig:sivewmel2}f pathway will become triply resonant and possibly contribute greater than the other pathways. 
Such instances are system dependent. 
For consistency, discussion will be isolated to the \autoref{fig:sivewmel2}a pathway in the sections that follow.

@@

\autoref{ABterms_DR} shows important selection rules for the vibronic transitions.
The $A$ term is allowed whenever an electronic transition has a one- and two- photon transition moment.
The brightness is controlled by vibrational overlap (Franck-Condon factors).
For centrosymmetric molecules, however, an electronic transition cannot be both one- and two- photon allowed, so the $A$ term will vanish.\cite{Milojevich2013, RN230}
This makes HDFG a unique tool for studying the electronic structure of centrosymmetric species in isotropic media, as it is sensitive to non-Condon effects.\cite{Olson2018}
In the absence of an $A$ term, the $B$ terms will dominate.
The $B_1$ and $B_2$ terms involve different types of vibronic coupling pathways.
In $B_1$, the Herzberg-Teller coupling is associated with the two-photon absorption process, whereas in $B_2$, the Herzberg-Teller coupling is associated with one-photon emission.
As such, $B_1$ contributions to a HDFG spectrum will be sensitive to vibronic coupling between the ground state of a mode and vibrational states on the $|e)$ surface. 
$B_2$ contributions are sensitive to vibronic coupling between the first excited state of a mode $\ket{g,1}$ and vibrational states on $|e)$.

@@

 where N$_\text{bulk}$ is the bulk number density ($\sim$ 10$^{28}$ m$^{-3}$ for H$_2$O$_{(l)}$ at room temperature), and N$_\text{surf}$ is the surface number density ($\sim$ 10$^{16}$ m$^{-2}$ for H$_2$O adsorbed at quartz).\cite{Du1994}	
 For simplicity, we assume that $\chi^{(2)}$ is a spatial constant so that $\int_{0^-}^\ell \mathrm{d}z \langle \alpha_{ij}\mu_{k} \rangle \approx \ell \langle \alpha_{ij}\mu_{k} \rangle$, where $z$ is oriented along the surface normal and $0^{-}$ is the edge of the interface where substrate begins. \cite{Su1998}
 We take the film thickness ($\ell$) probed by vSFG to be 1 nm, much less than the vSFG coherence length.\cite{RN133}
 A normal dispersion curve is assumed so that $F(\omega_1+\omega_2) \approx F(2\omega_2-\omega_1)$.
 To simplify analysis, orientational averaging is ignored (e.g., $\langle \beta_{ijk} \mu_{l} \rangle \approx \beta \mu$) and the input and output fields are taken to be co-polarized, so that
 \begin{equation}
	 		P_\text{ratio} \equiv \frac{\abs{P^{(3)}_{\text{HDFG}}}}{\abs{P^{(2)}_{\text{vSFG}}}} \approx \frac{N_\text{bulk}}{N_\text{surf}} \frac{\beta}{\alpha} E(\omega_2) \ell \sim 10^3 \frac{\beta}{\alpha} E(\omega_2)\\
	 \end{equation}
 Ziegler has noted that for a field with intensity 10 GW/cm$^{2}$, $\frac{\beta E}{\alpha} \sim 10^{-3} $ for vibrational modes when $E(\omega_2)$ is largely detuned from electronic resonances. \cite{RN515}
 Such an intensity is easily obtained using modern ultrafast sources.
 In this limit, $P_\text{ratio} \sim 1$.
 Since the intensity ratio scales as $\abs{P_\text{ratio}}^2$, we see that the HDFG output is roughly as strong as vSFG, assuming only interfacial contributions in vSFG.
 It is important to note that this calculation assumed negligible dipole and Raman polarizability dependence upon the distance from surface normal, ignored orientational averaging effects, and presumed equivalence of local field factors, all of which can significantly reduce output from either process. 
 Nevertheless, since HDFG produces a number of photons similar to that of vSFG, HDFG is a viable technique for investigating bulk systems.
 This method provides practitioners of sum frequency generation the ability to probe isotropic dephasing dynamics without many meaningful changes in their optical setup, and thus compare  dephasing in the bulk and at the interface.\cite{RN224}

>>>>

\section{Introduction}
Coherent multidimensional spectroscopy (CMDS) is a family of nonlinear spectroscopy methods that form the optical analogue of multidimensional nuclear magnetic resonance (NMR) spectroscopy.\cite{Cho2008}
Multiresonant, four wave mixing CMDS experiments, first proposed by Oudar and Shen in 1980,\cite{RN307} directly probe coupling and correlations between different vibrational, electronic, and vibronic states. \cite{RN281, RN103, Cho2008} 
Four wave mixing CMDS has resolved anharmonicities, ultrafast dynamics, and other inter- and intra-molecular couplings in numerous systems. \cite{Cho2008, Gaynor2017, Ziegler2018, Ogilvie2019, Bonn2021, RN325, Chen2024}
To target vibrational and electronic coupling, methods commonly make use of Raman transitions, e.g., doubly vibrationally enhanced spectroscopy (DOVE), triple sum frequency (TSF), coherent anti-Stokes Raman spectroscopy (CARS), resonant impulse stimulated Raman spectroscopy (RISRS).\cite{RN103, Dhar1994, RN345, Cho2000, RN491}
Here we explore the less-utilized hyper-Raman transitions in four wave mixing CMDS.

@@

\autoref{fig:comparisonwmel} shows both HDFG and HSFG processes and compares them with related spectroscopies that investigate vibrational-electronic coupling.
The three-wave mixing ($\chi^{(2)}$) techniques are often less applicable as they require non-centrosymmetry.\cite{RN227}
Hyper-Raman scattering itself is a six-wave mixing ($\chi^{(5)}$) technique; stimulated hyper-Raman will be weak and susceptible to FWM cascades.\cite{RN515, RN243, Cho2000_Cascade}
A coherent six-wave mixing analogue of hyper-Raman scattering, coherent anti-Stokes hyper-Raman spectroscopy (CAHRS), has been proposed and explored theoretically by several authors; however, it is also highly susceptible to FWM cascades.\cite{Berger1978, Bjarnason1980, Cho1997, Cho1998}
Of the many four-wave mixing techniques, HDFG and HSFG have strong similarity to Raman, stimulated Raman, and RISRS in that the electronic state must remove or add a $v$ quantum to the ground state (``ev'' coupling). %rpm: not sure what you're trying to say by this
With the other four-wave mixing techniques (\autoref{fig:comparisonwmel}, second row), the output signal depends on more states than just a single vibrational mode and electronic state (``evx'' coupling). 
As a result, the relationship between the output and vibrational-electronic coupling is more detailed, but also more complicated.
HDFG, HSFG and Raman are more direct probes of electronic-vibrational coupling.
Notably, HDFG and HSFG can be implemented as 2-color experiments (with $\omega_2=\omega_3$);\cite{Cho2001} vibrational sum frequency generation (vSFG) or difference frequency generation (vDFG) setups can be trivially reconfigured to perform HSFG or HDFG measurements, respectively, to investigate ev coupling in bulk systems.

@@

Inspired by recent work that demonstrated the presence of HDFG in ultrafast DOVE and terahertz-IR-visible (TIRV) experiments, we investigate the parameters which result in HDFG output. \cite{Bonn2024, McDonnell2024}
This paper is organized as follows.
In \autoref{steadystate}, we first identify the selection rules that drives HDFG in the absence of electronic resonance. 
The selection rules of electronically-resonant HDFG are then identified through a Herzberg-Teller dipole expansion.
HDFG is found to be allowed in any harmonic system that has infrared active vibrations.
A simple harmonic oscillator model system is used to simulate HDFG spectra and illustrate how the excited state potential energy surface affects the hyper-Raman excitation spectrum.
After developing selection rules, the site selective properties, useful experimental aspects of HDFG, and a new scheme to extract the hyper-Raman hyperpolarizability ($\beta_{ijk}$) from HDFG spectra are discussed in \autoref{quant}.
The main findings are summarized in \autoref{conclusion}.

@@

It is useful to expose relationships between nonlinear output, transition dipoles and hyper-Raman hyperpolarizabilities in the driven limit. \cite{Simpson2004}
Under the electric dipole approximation, the I$^{th}$ component of the third order nonlinear output polarization, ${P}^{(3)}_I$, of a four wave mixing process, induced by electric fields $E_J$, $E_K$, and $E_L$ at output frequency $\omega_4=-\omega_1 + \omega_2 + \omega_3$ is written as (using Einstein summation) \cite{RN307}
\begin{equation} \label{polarization}
{P}^{(3)}_I (\omega_4)  = \chi^{(3)}_{IJKL} E_J(\omega_3) E_K(\omega_2) E_L(\omega_1) 
\end{equation}
where $\chi^{(3)}_{IJKL}$ is the IJKL element of the third order electrical susceptibility, a rank four tensor, generally written as
\begin{equation}
	\chi^{(3)}_{IJKL} = NF(\omega_4) \langle \gamma_{ijkl} \rangle
\end{equation}
where $N$ is a number density, $F$ is the Lorentz local field factor, and $\gamma_{ijkl}$ is the third order polarizability (i.e., second hyperpolarizability). 
The brackets indicate an orientational (or rotational) average.\cite{Andrews1977}
Uppercase letters refer to laboratory frame coordinates and lower case letters refer to corresponding molecular frame coordinates.

@@

When $\omega_2 + \omega_3$ become electronically resonant, the interference changes for the \autoref{fig:sivewmel2}c-f pathways.
Note that only pathways (c,e,f) possess identical transition moments.
As a result, pathways (c,d) survive and can provide output when $\omega_2+\omega_3$ is resonant with an electronic state. 
These results provide a simple rule of thumb for collecting HDFG spectra: use \autoref{fig:sivewmel2}a,b pathways to measure spectra when significantly detuned from an electronic resonance, and any pathway can measure spectra when electronically resonant.
However, since electronic states commonly have short dephasing times and have significant inhomogeneous broadening,\cite{Dong2015, Lewis2015} it is likely the \autoref{fig:sivewmel2}f pathway will become triply resonant and possibly contribute greater than the other pathways. 
Such instances are system dependent. 
For consistency, discussion will be isolated to the \autoref{fig:sivewmel2}a pathway in the sections that follow.

@@

\autoref{ABterms_DR} shows important selection rules for the vibronic transitions.
The $A$ term is allowed whenever an electronic transition has a one- and two- photon transition moment.
The brightness is controlled by vibrational overlap (Franck-Condon factors).
For centrosymmetric molecules, however, an electronic transition cannot be both one- and two- photon allowed, so the $A$ term will vanish.\cite{Milojevich2013}
This makes HDFG a unique tool for studying the electronic structure of centrosymmetric species in isotropic media, as it is sensitive to non-Condon effects.\cite{Olson2018}
In the absence of an $A$ term, the $B$ terms will dominate.
The $B_1$ and $B_2$ terms involve different types of vibronic coupling pathways.
In $B_1$, the Herzberg-Teller coupling is associated with the two-photon absorption process, whereas in $B_2$, the Herzberg-Teller coupling is associated with one-photon emission.
As such, $B_1$ contributions to a HDFG spectrum will be sensitive to vibronic coupling between the ground state of a mode and vibrational states on the $|e)$ surface. 
$B_2$ contributions are sensitive to vibronic coupling between the first excited state of a mode $\ket{g,1}$ and vibrational states on $|e)$.

@@

 where N$_\text{bulk}$ is the bulk number density ($\sim$ 10$^{28}$ m$^{-3}$ for H$_2$O$_{(l)}$ at room temperature), and N$_\text{surf}$ is the surface number density ($\sim$ 10$^{16}$ m$^{-2}$ for H$_2$O adsorbed at quartz).\cite{Du1994}	
 For simplicity, we assume that $\chi^{(2)}$ is a spatial constant so that $\int_{0^-}^\ell \mathrm{d}z \langle \alpha_{ij}\mu_{k} \rangle \approx \ell \langle \alpha_{ij}\mu_{k} \rangle$, where $z$ is oriented along the surface normal and $0^{-}$ is the edge of the interface where substrate begins. 
 We take the film thickness ($\ell$) probed by vSFG to be 1 nm, much less than the vSFG coherence length.\cite{RN133}
 A normal dispersion curve is assumed so that $F(\omega_1+\omega_2) \approx F(2\omega_2-\omega_1)$.
 To simplify analysis, orientational averaging is ignored (e.g., $\langle \beta_{ijk} \mu_{l} \rangle \approx \beta \mu$) and the input and output fields are taken to be co-polarized, so that
 \begin{equation}
	 		P_\text{ratio} \equiv \frac{\abs{P^{(3)}_{\text{HDFG}}}}{\abs{P^{(2)}_{\text{vSFG}}}} \approx \frac{N_\text{bulk}}{N_\text{surf}} \frac{\beta}{\alpha} E(\omega_2) \ell \sim 10^3 \frac{\beta}{\alpha} E(\omega_2)\\
	 \end{equation}
 Ziegler has noted that for a field with intensity 10 GW/cm$^{2}$, $\frac{\beta E}{\alpha} \sim 10^{-3} $ for vibrational modes when $E(\omega_2)$ is largely detuned from electronic resonances. \cite{RN515}
 Such an intensity is easily obtained using modern ultrafast sources.
 In this limit, $P_\text{ratio} \sim 1$.
 Since the intensity ratio scales as $\abs{P_\text{ratio}}^2$, we see that the HDFG output is roughly as strong as vSFG, assuming only interfacial contributions in vSFG.
 It is important to note that this calculation assumed negligible dipole and Raman polarizability dependence upon the distance from surface normal, ignored orientational averaging effects, and presumed equivalence of local field factors, all of which can significantly reduce output from either process. 
 Nevertheless, since HDFG produces a number of photons similar to that of vSFG, HDFG is a viable technique for investigating bulk systems.
 This method provides practitioners of sum frequency generation the ability to probe isotropic dephasing dynamics without many meaningful changes in their optical setup, and thus compare  dephasing in the bulk and at the interface.\cite{RN224}
